---
title: "The Effect of Education on Income: A Cross-Country Analysis"
author: "Sabin Cyriac"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(WDI)
library(dplyr)
```

## Introduction

Education is often considered a key factor in economic development. In this project, we investigate whether countries with higher average levels of education also tend to have higher income, measured by GDP per capita.

**Research Question**:  
Does higher education lead to higher income at the country level?

---

## Methodology

We use cross-country data from the World Bank including:
- **GDP per capita**: Proxy for income
- **Average years of schooling**: Proxy for education
- **Life expectancy**: Health indicator (control)
- **Labor force participation rate**: Proxy for employment
- **Inflation rate**: Control for macroeconomic stability

We will clean and analyze this data using R.


# Data Collection


```{r load_data_2023}
library(WDI)
library(dplyr)

# Download World Bank indicators for 2023
education_income_data <- WDI(
  country = "all",
  indicator = c(
    "NY.GDP.PCAP.CD",     # GDP per capita
    "SE.SCH.LIFE",        # Average years of schooling
    "SP.DYN.LE00.IN",     # Life expectancy
    "SL.TLF.CACT.ZS",     # Labor force participation
    "FP.CPI.TOTL.ZG"      # Inflation
  ),
  start = 2023,
  end = 2023,
  extra = TRUE
)

# Rename columns
education_income_data <- education_income_data %>%
  rename(
    gdp_per_capita = NY.GDP.PCAP.CD,
    school_years = SE.SCH.LIFE,
    life_expectancy = SP.DYN.LE00.IN,
    labor_force = SL.TLF.CACT.ZS,
    inflation = FP.CPI.TOTL.ZG
  )

# Select only relevant columns and remove aggregates
education_income_data <- education_income_data %>%
  select(country, year, region, income, gdp_per_capita, school_years, life_expectancy, labor_force, inflation) %>%
  filter(region != "Aggregates")

# View first 10 rows
knitr::kable(head(education_income_data, 10))



```


## Data Cleaning


```{r clean_data}

# Remove aggregate regions and missing values
education_income_data_2023 <- education_income_data_2023 %>%
  filter(region != "Aggregates", !is.na(SP.DYN.LE00.IN))


# Remove duplicates, just in case
education_income_data <- education_income_data %>% distinct()

# Preview cleaned data
head(education_income_data)


```
```{r removing empty rows}
# Load ggplot2 if not already loaded
library(ggplot2)

#calculate inflation per region

inflation_by_region <- education_income_data %>%
  filter(!is.na(inflation), !is.na(region)) %>%
  group_by(region) %>%
  summarise(avg_inflation = mean(inflation, na.rm = TRUE)) %>%
  arrange(desc(avg_inflation))

ggplot(inflation_by_region, aes(x = reorder(region, avg_inflation), y = avg_inflation)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Average Inflation Rate by Region (2023)",
    x = "Region",
    y = "Average Inflation Rate (%)"
  ) +
  theme_minimal()

# Prepare data: calculate average life expectancy per region
life_expectancy_by_region <- education_income_data %>%
  group_by(region) %>%
  summarise(avg_life_expectancy = mean(life_expectancy, na.rm = TRUE)) %>%
  arrange(desc(avg_life_expectancy))

ggplot(life_expectancy_by_region, aes(x = reorder(region, avg_life_expectancy), y = avg_life_expectancy)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  coord_flip() +
  labs(
    title = "Average Life Expectancy by Region (2023)",
    x = "Region",
    y = "Average Life Expectancy (Years)"
  ) +
  theme_minimal()

# Calculate average GDP per capita by region
gdp_by_region <- education_income_data %>%
  group_by(region) %>%
  summarise(avg_gdp = mean(gdp_per_capita, na.rm = TRUE)) %>%
  arrange(desc(avg_gdp))

ggplot(gdp_by_region, aes(x = reorder(region, avg_gdp), y = avg_gdp)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  labs(
    title = "Average GDP per Capita by Region (2023)",
    x = "Region",
    y = "GDP per Capita (USD)"
  ) +
  theme_minimal()



# Visualize average years of schooling by region
schooling_by_region <- education_income_data %>%
  filter(!is.na(school_years), !is.na(region)) %>%
  group_by(region) %>%
  summarise(avg_schooling_years = mean(school_years, na.rm = TRUE)) %>%
  arrange(desc(avg_schooling_years))

ggplot(schooling_by_region, aes(x = reorder(region, avg_schooling_years), y = avg_schooling_years)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average Years of Schooling by Region",
    x = "Region",
    y = "Average Years of Schooling"
  ) +
  theme_minimal()



```
 


```{r regression_model}


# Only run diagnostics if model was created
if (exists("regression_model")) {
  par(mfrow = c(2, 2))  # Arrange plots in 2x2 grid
  plot(regression_model)
  par(mfrow = c(1, 1))  # Reset layout
} else {
  cat("Regression model not found. Please ensure regression was run.")
}

# Descriptive summary statistics for key variables

summary_stats <- education_income_data %>%
  summarise(
    avg_gdp = mean(gdp_per_capita, na.rm = TRUE),
    avg_school = mean(school_years, na.rm = TRUE),
    avg_life = mean(life_expectancy, na.rm = TRUE),
    avg_inflation = mean(inflation, na.rm = TRUE),
    avg_labor = mean(labor_force, na.rm = TRUE)
  )

# View the summary statistics table
knitr::kable(summary_stats)



# Calculate average GDP, schooling, life expectancy, inflation, labor force by region
region_comparisons <- education_income_data %>%
  group_by(region) %>%
  summarise(
    avg_gdp = mean(gdp_per_capita, na.rm = TRUE),
    avg_schooling_years = mean(school_years, na.rm = TRUE),
    avg_life_expectancy = mean(life_expectancy, na.rm = TRUE),
    avg_inflation = mean(inflation, na.rm = TRUE),
    avg_labor_force = mean(labor_force, na.rm = TRUE)
  )

# View the results
knitr::kable(region_comparisons)

# Descriptive summary statistics for key variables
summary_stats <- education_income_data %>%
  summarise(
    avg_gdp = mean(gdp_per_capita, na.rm = TRUE),
    avg_school = mean(school_years, na.rm = TRUE),
    avg_life = mean(life_expectancy, na.rm = TRUE),
    avg_inflation = mean(inflation, na.rm = TRUE),
    avg_labor = mean(labor_force, na.rm = TRUE)
  )



# Create a data frame with only the relevant variables for correlation
cor_data <- education_income_data %>%
  select(gdp_per_capita, school_years, life_expectancy, labor_force, inflation)

# Compute the correlation matrix using pairwise.complete.obs to handle missing data
cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

# Show the correlation matrix
knitr::kable(cor_matrix)


```
```{r}

# Findings

The analysis shows that regions with higher average years of schooling tend to have higher GDP per capita. This supports the hypothesis that education contributes to economic development. Life expectancy and labor force participation also appear to play roles, suggesting a broader relationship between health, workforce engagement, and income.
Inflation varies across regions but does not show a clear direct relationship with GDP in this analysis. A formal regression model could not be estimated due to missing data, which limits the ability to draw causal inferences.
Still, the visualizations and correlation matrix provide valuable insight into patterns across countries.

# Conclusion

This project investigated the relationship between education and income across countries using World Bank data.Although I encountered with missing data that prevented full regression modeling, the descriptive analysis and correlation patterns suggest that countries with more schooling tend to enjoy higher incomes. Future research could expand on this using more complete longitudinal data and additional controls.This project explored the relationship between education and income across countries using 2023 World Bank data.Eventhough challenges with missing values that limited the ability to perform a full regression analysis, the visualizations and descriptive statistics provided useful insights.

The findings suggest that countries with higher average years of schooling generally tend to have higher GDP per capita. This aligns with economic theory, which holds that education improves human capital and productivity. Additionally, regions with higher life expectancy and labor force participation also showed relatively stronger economic indicators, suggesting that these factors may interact with education to influence income levels.

Due to data limitations—particularly missing values for key variables in many countries—a full multivariable regression model could not be estimated. This represents a limitation in the analysis, and it emphasizes the importance of complete datasets for robust statistical modeling.



Limitations include missing values, single-year data (2023), and potential unobserved variables. Despite this, the analysis highlights the importance of education in global economic outcomes.


# References

- World Bank Indicators (WDI). [https://data.worldbank.org](https://data.worldbank.org)
- OECD Education Database. [https://data.oecd.org/education.htm](https://data.oecd.org/education.htm)


